Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-21 submission/manuscript.Rmd
add references for scfa gene specific look
update methods write up
add figure legend template write up

**Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors. These genes were based on pathways from KEGG and can be found in the supplemental material [Table S1].

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. Models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002]. Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. The measured SCFA concentrations were first tested for differences between groups. The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.

@@

Insert figure legends with the first sentence in bold, for example:

**Figure 1. Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B). Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).

@@

>>>>

**Pulling genes involved with SCFA synthesis.** Specific genes located near the end of the pathways involved in the synthesis of acetate, butyrate, and propionate were analyzed for any differences between individuals with normal colons and those with tumors. These genes were based on pathways from KEGG as well as previous research [@scfa_baxter2014; @Kanehisa2015] and a list can be found in the supplemental material [Table S1].

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification. Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002]. Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. The measured SCFA concentrations were first tested for differences between groups. The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.

@@

**Figure 1. Using HPLC no change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A). Butyrate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (B). Propionate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (C). The black links indicate the median SCFA concentration. Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D). Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F). The black dots and lines represent the median change in SCFA concentration.

**Figure 2. SCFAs do not improve OTU-based Random Forest models.** Difference between the area under the curve of 100 different 80/20 split OTU-based normal versus adenoma 10-fold CV models with and without SCFAs (A). Difference between the area under the curve of 100 different 80/20 OTU-based normal versus carcinoma 10-fold CV models with and without SCFAs (B). The black linke represents the median AUC. The dotted line highlights an AUC of 0.5.

**Figure 3. No change in butyrate producing genes identified between normal, adenoma, and carcinoma individuals.** Imputed gene relative abundance of important butyrate pathway genes using PICRUSt (A). Counts per million (corrected for size and number of contigs in an OPF) for the Butyrate Kinase gene (B). The other genes from the PICRUSt analysis did not align to any of the OPFs in the metagenome analysis.

@@

**Figure S1. OTU-based Random Forest models of SCFA concentrations.** Classification Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on higher or lower than the medain SCFA concentration (A). The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (B). Regression Random Forest train and tests of 100 different 80/20 OTU-based models with 10-fold CV based on correlation to actual SCFA concentration (C). The top 10 OTUs based on mean decrease in accuracy (MDA) for each model, colored by their lowest taxonomic identification (D).

### 2018-05-21 submission/manuscript.tex
update methods write up
add figure legend template write up

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (13,
14) and can be found in the supplemental material {[}Table S1{]}.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (15). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. Models to classify high or low SCFA concentration
based on 16S rRNA gene sequencing data were created using Random Forest
(16). Regression models to classify the exact SCFA concentration based
on 16S rRNA gene sequencing data also were built using the Random Forest
algorithm. The measured SCFA concentrations were first tested for
differences between groups. The ability of 16S rRNA gene sequencing to
classify these concentrations were then assessed. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed.

@@

Insert figure legends with the first sentence in bold, for example:

\textbf{Figure 1. Number of OTUs sampled among bacterial and archaeal
16S rRNA gene sequences for different OTU definitions and level of
sequencing effort.} Rarefaction curves for different OTU definitions of
Bacteria (A) and Archaea (B). Rarefaction curves for the coarse
environments in Table 1 for Bacteria (C) and Archaea (D).

@@

>>>>

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (13,
14) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (15). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. We next assessed whether SCFAs added to Random
Forest models built with OTU data significantly improved classification.
Next, models to classify high or low SCFA concentration based on 16S
rRNA gene sequencing data were created using Random Forest (16).
Regression models to classify the exact SCFA concentration based on 16S
rRNA gene sequencing data also were built using the Random Forest
algorithm. The measured SCFA concentrations were first tested for
differences between groups. The ability of 16S rRNA gene sequencing to
classify these concentrations were then assessed. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed.

@@

\textbf{Figure 1. Using HPLC no change in SCFA measurements was observed
between normal, adenoma, and carcinoma individuals.} Acetate
concentrations in fecal samples of individuals with normal colons,
adenomas, and carcinomas (A). Butyrate concentrations in fecal samples
of individuals with normal colons, adenomas, and carcinomas (B).
Propionate concentrations in fecal samples of individuals with normal
colons, adenomas, and carcinomas (C). The black links indicate the
median SCFA concentration. Acetate concentrations in fecal samples
before and after treatment for adenoma (yellow) and carcinoma (red) (D).
Butyrate concentrations in fecal samples before and after treatment for
adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in
fecal samples before and after treatment for adenoma (yellow) and
carcinoma (red) (F). The black dots and lines represent the median
change in SCFA concentration.

\textbf{Figure 2. SCFAs do not improve OTU-based Random Forest models.}
Difference between the area under the curve of 100 different 80/20 split
OTU-based normal versus adenoma 10-fold CV models with and without SCFAs
(A). Difference between the area under the curve of 100 different 80/20
OTU-based normal versus carcinoma 10-fold CV models with and without
SCFAs (B). The black linke represents the median AUC. The dotted line
highlights an AUC of 0.5.

\textbf{Figure 3. No change in butyrate producing genes identified
between normal, adenoma, and carcinoma individuals.} Imputed gene
relative abundance of important butyrate pathway genes using PICRUSt
(A). Counts per million (corrected for size and number of contigs in an
OPF) for the Butyrate Kinase gene (B). The other genes from the PICRUSt
analysis did not align to any of the OPFs in the metagenome analysis.

@@

\textbf{Figure S1. OTU-based Random Forest models of SCFA
concentrations.} Classification Random Forest train and tests of 100
different 80/20 OTU-based models with 10-fold CV based on higher or
lower than the medain SCFA concentration (A). The top 10 OTUs based on
mean decrease in accuracy (MDA) for each model, colored by their lowest
taxonomic identification (B). Regression Random Forest train and tests
of 100 different 80/20 OTU-based models with 10-fold CV based on
correlation to actual SCFA concentration (C). The top 10 OTUs based on
mean decrease in accuracy (MDA) for each model, colored by their lowest
taxonomic identification (D).
